\documentclass[../TGF.tex]{subfiles}

\begin{document}
\section{Introduction}
\subsection{Motivation}
\label{sec:motivation}
We begin with an observation: functions found in computer programming may not
behave like functions in mathematics.

Unfortunately, it is common to call computer functions and mathematical
functions both functions. In this report we will do so, as it is usually clear
by the context, and only specifying when there might be confusion. In these
situations, we may refer to a computer function as a~\emph{procedure}.

An example of this is the \mintinline{C}{malloc()} function found in
the~\mintinline{C}{stdlib.h} library in the~\mintinline{C}{C} programming
language, which takes a size in number of bits as input and outputs an unique
pointer to an address to a block of memory allocated to fit the requested size,
which is marked as reserved. This function

\begin{enumerate}
    \item generally outputs a different pointer on calls with the same input,
    \item modifies global state by marking allocated memory as reserved.
\end{enumerate}

These properties are incompatible with the definition of a mathematical
function. The first one means it is not well defined, and the second one means
it has side effects, which mathematical functions do not have. This motivates
the following definitions:

\begin{definition}[Pure function]
    We say that a procedure that
    \begin{enumerate}
        \item Outputs the same result for the same inputs (\ie is well defined),

        \item Does not modify global state.
    \end{enumerate}
    is~\emph{pure}.
    Conversely, we say a procedure that is not pure is~\emph{impure}.
\end{definition}

So, the~\mintinline{C}{malloc()} function is impure, and a function that outputs
the result of squaring an integer is pure.

In theory, a compiler could use its knowledge of certain functions being pure to
better reason about code, which can be used to obtain benefits such as

\begin{enumerate}
    \item If a pure function is called more than once with the
        same input, the result calculated on the previous run can be reused
        without having to call the function again~\cite{frostMemorization}. This
        is called~\emph{memorization}

    \item Since pure functions have no side-effects, it is usually safe to call
        them in parallel \cite{SussPureFunctionParallelization}, leading to
        great optimization.
\end{enumerate}

While pure functions present some significant benefits, they also come with some
apparent drawbacks. For example, it is not obvious how pure functions could
react to user input or output anything to the screen, such as debugging
information or logs. They also seem to be limited to not using global variables.

Most languages circumvent these problems by providing an interface to define
both pure and impure functions. In this way, some of the benefits can be
applied, and the drawbacks can be handled by impure functions.

As of this point, the problems we mentioned are only apparent, and while it is
not at all obvious how they could be avoided, there is also no definitive proof
they cannot. We promise to provide an abstraction that solves these limitations,
and then we will see how it can be successfully implemented into a useful
programming language.

\subsection{Overview}
The idea of using mathematical functions to model procedures is the starting
point of~\emph{categorical semantics} of programming. Good semantics (meaning
sound and complete) allows to reason about correctness of programs, as well as
termination and complexity. The naive view on procedures suggests that the
semantics takes value in the category of sets. The subtleties regarding the
notions of procedures, and in particular impure procedures in programming, make
this problematic.

The big discovery by Moggi in~\emph{Notions of computation and
monads}~\cite{Moggi-notions-computation-monads} 1991, is that functions that are
impure from the viewpoint of the category of sets may become pure in other
categories, and that these other categories are~\emph{Kleisli categories for
monads}. There is one monad for each computational effect to achieve. The most
important of these effects are IO, exceptions, state, non-determinism, and so
on, the corresponding monads of which are well-known mathematical structures.
While it was a big theoretic advance to be able to find categorical semantics in
Kleisli categories, perhaps Moggi had not predicted how important it would be
also in practical programming.

This only became clear with the development of Haskell, one of the first
programming languages to actually take these ideas in as a foundational
principle. Since Haskell is based on category theory and designed to work well
with monads, it can be more pure than traditional programming languages. With
time, monadic ideas, and hence more principled programming, have been adopted in
some form by most mainstream programming languages. Not as a foundation, but as
an optional tool.

This is a huge area of research, and it has changed the whole panorama of
theoretical computer science, and is also impacting more mainstream programming.
The goal of this bachelor's thesis is modest, just to explain the basic
principles of how monads are used to model effects.

In Section 1 we gave some motivation and context to our problem.
Then in Section 2 we introduce the basic categorical notions we will build on to
explain these results. Section 3 contains a brief introduction to the results
Moggi showed in his famous paper, and in Section 4 we finish the mathematical
part of the work with a treatment of the relationship between adjunctions and
monads. We close with Section 5, where we start by giving a concise introduction
to Haskell and its notation, to then explore how these abstractions are
successfully implemented and used in practice.

\subsection{Acknowledgements}
I would like to express my sincere thank Joachim Kock and Thomas Mikhail, my
supervisors for this bachelor's thesis, for the help they provided and their
involvement from day one.
\end{document}
